---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r echo=TRUE}
options(scipen=1, digits=2)
library(dplyr)
library(ggplot2)

unzip("activity.zip")
data <- read.csv("activity.csv", as.is = TRUE)
data$date <- as.Date(data$date)
data.no.na <- data[!is.na(data$steps),]
```

## What is mean total number of steps taken per day?
```{r echo=TRUE}
steps.by.date <- data.no.na %>% group_by(date) %>% summarise(total.steps = sum(steps))
ggplot(steps.by.date, aes(total.steps)) + geom_histogram() + ggtitle("Histogram of the Total Number of Steps Taken Each Day")

steps.mean <- mean(steps.by.date$total.steps, na.rm = TRUE)
steps.median <- median(steps.by.date$total.steps, na.rm = TRUE)
```
Mean total number of steps taken per day is `r steps.mean`.  
Median total number of steps taken per day is `r steps.median`.  

## What is the average daily activity pattern?

```{r echo=TRUE}
steps.by.interval <- data.no.na %>% group_by(interval) %>% summarise(mean.steps = mean(steps))
ggplot(steps.by.interval, aes(interval, mean.steps)) + geom_line() + ggtitle("Average Daily Activity Pattern")

max.interval <- steps.by.interval$interval[which.max(steps.by.interval$mean.steps)]
```
The 5-minute interval `r max.interval`, on average across all the days in the dataset, contains the maximum number of steps.

## Imputing missing values

Total number of missing values in the dataset is `r sum(is.na(data$steps))`.
We fill in the missing values by the mean for that 5-minute interval.

```{r echo=TRUE}
na.idx <- is.na(data$steps)
data.na <- data[na.idx,]
data.na.joined <- data.na %>% left_join(steps.by.interval, by="interval")
data.na.filled <- data
data.na.filled[na.idx, "steps"] <- data.na.joined$mean.steps

steps.by.date.filled <- data.na.filled %>% group_by(date) %>% summarise(total.steps = sum(steps))
ggplot(steps.by.date.filled, aes(total.steps)) + geom_histogram() + ggtitle("Histogram of the Total Number of Steps Taken Each Day")

steps.mean.filled <- mean(steps.by.date.filled$total.steps, na.rm = TRUE)
steps.median.filled <- median(steps.by.date.filled$total.steps, na.rm = TRUE)
```
Mean total number of steps taken per day is `r steps.mean.filled`.  
Median total number of steps taken per day is `r steps.median.filled`.

Imputing missing data increases the median to be closer to the mean. It also increases the daily number of steps for those days with some NA steps originally.

## Are there differences in activity patterns between weekdays and weekends?

```{r echo=TRUE}
data2 <- data.no.na
data2$dow <- weekdays(data2$date)
data2$weekday.or.weekend <- if_else(data2$dow %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
steps.by.interval.2 <- data2 %>% group_by(weekday.or.weekend, interval) %>% summarise(mean.steps=mean(steps))

ggplot(steps.by.interval.2, aes(interval, mean.steps)) + geom_line() + ggtitle("Average Daily Activity Pattern") + facet_grid(weekday.or.weekend ~ .)
```

